---
title: "DACSS-603_Spring25:Homework 1"
author: "Kuan-Cheng Lee"
format: 
  html: 
    embed-resources: true
    self-contained-math: true
---

#library installation
```{r library}
library(ggplot2)
library(dplyr)

```
# Question 1

First, let's download the LungCap.csv file and save it to your Quarto project folder. Read in the data file. 

```{r, echo=T}
df <- read.csv("LungCapData.csv")
```


The distribution of LungCap looks as follows:

```{r, echo=T}
hist(df$LungCap)
```

The histogram suggests that the distribution is close to a normal distribution. Most of the observations are close to the mean. Very few observations are close to the margins (0 and 15).

a. distribution of LungCup:

```{r ,echo=T}

ggplot(df, aes(x = LungCap)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Probability Density of Lung Capacity", x = "Lung Capacity", y = "Density") +
  theme_minimal()



```


b. Comparing LungCap Distribution by Gender:

```{r ,echo=T}

ggplot(df, aes(x = Gender, y = LungCap, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Lung Capacity by Gender", x = "Gender", y = "Lung Capacity") +
  theme_minimal()


```


c. Comparing Mean LungCap for Smokers vs. Non-Smokers:
```{r ,echo=T}

df %>%
  group_by(Smoke) %>%
  summarize(mean_LungCap = mean(LungCap, na.rm = TRUE),
            sd_LungCap = sd(LungCap, na.rm = TRUE),
            n = n())


```


d. Relationship Between Smoking and Lung Capacity Across Age Groups:
```{r ,echo=T}

df <- df %>%
  mutate(AgeGroup = case_when(
    Age <= 13 ~ "≤13",
    Age >= 14 & Age <= 15 ~ "14-15",
    Age >= 16 & Age <= 17 ~ "16-17",
    Age >= 18 ~ "≥18"
  ))
ggplot(df, aes(x = AgeGroup, y = LungCap, fill = Smoke)) +
  geom_boxplot() +
  labs(title = "Lung Capacity by Age Group and Smoking Status", x = "Age Group", y = "Lung Capacity") +
  theme_minimal()


```


e. Comparing LungCap for Smokers and Non-Smokers in Each Age Group: 

```{r ,echo=T}
df %>%
  group_by(AgeGroup, Smoke) %>%
  summarize(mean_LungCap = mean(LungCap, na.rm = TRUE),
            sd_LungCap = sd(LungCap, na.rm = TRUE),
            n = n())


```


# Question 2
Setting up the database
```{r ,echo=T}
total_prisoners <- 920
X <- c(0,1,2,3,4)
Frequency <- c(150, 420, 180, 100, 70)

prob <- Frequency/ total_prisoners

data <- data.frame(X, Frequency, prob)
print(data)


```

What is the probability that a randomly selected inmate has exactly 2 prior convictions?
X = 2   **The answer is 0.1956522**
```{r ,echo=T}
p_2 <- prob[X == 2]
print(p_2)

```

What is the probability that a randomly selected inmate has fewer than 2 prior convictions?
X < 2   **The answer is 0.6195652**
```{r ,echo=T}
p_less_than_2 <- sum(prob[X < 2])

print(p_less_than_2)

```

What is the probability that a randomly selected inmate has 2 or fewer than 2 prior convictions?
X < = 2   **The answer is 0.8152174**
```{r ,echo=T}
p_less_than_or_equal_2 <- sum(prob[X <= 2]) #P(X = 0) + P(X = 1) + P(X = 2)
print(p_less_than_or_equal_2)


```

What is the probability that a randomly selected inmate has more than 2 prior convictions?
X > 2   **The answer is 0.1847826**
```{r ,echo=T}
p_more_than_2 <- sum(prob[X > 2]) #P(X = 3) + P(X = 4)
print(p_more_than_2)

```

What is the expected value for the number of prior convictions?
**The value is 1.478261**
```{r ,echo=T}
expected_value <- sum(X * prob)
print(expected_value)

```

Calculte the variance and the standard deviation for the Prior Convictions.
**The varience is 1.249527. The standard deviation is 1.117823.**
```{r ,echo=T}
E_X2 <- sum((X^2) * prob) 
varience <- E_X2 - (expected_value^2)
std_dev <- sqrt(varience)

print(varience)
print(std_dev)

```




